---
title: "DNAcycP2: **DNA** **Cyc**lizability **P**rediction v**2**"
author: "Brody Kendall, Ji-Ping Wang, and Keren Li"
date: "`r Sys.Date()`"
output: 
    rmarkdown::html_document:
        highlight: pygments
        toc: true
        fig_width: 5
vignette: >
    %\VignetteIndexEntry{dnacycp2}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEncoding{UTF-8}
---

**Maintainer**: Ji-Ping Wang, <<jzwang@northwestern.edu>>


```{r setup, include = FALSE}
library(dnacycp2)
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>"
)
```


**References for methods**:

TODO: Update citation when applicable

****************************

**DNAcycP2**, short for **DNA** **cyc**lizablity **P**rediction v**2**, is an R 
package for accurate, unbiased prediction of DNA intrinsic cyclizablity score. 
It was built upon a deep learning architecture with a hybrid of Inception and 
Residual network structure and an LSTM layer. DNAcycP2 is an updated version of 
DNAcycP, released by Li et al 2021 (see below). DNAcycP was trained based on 
loop-seq data from Basu et al 2021 (see below), while DNAcycP2 was trained 
based on smoothed predictions of this loop-seq data. The predicted score (for 
either DNAcycP or DNAcycP2), termed **C-score** achieves high accuracy compared 
to the experimentally measured cyclizablity score by loop-seq assay.

## Key differences between DNAcycP2 and DNAcycP

Following the release of DNAcycP, it was discovered that the training data 
contained residual measurement bias, leading to biased predictions. To correct 
this bias in the data, we employed a data augmentation + periodic smoothing 
approach to generate new, unbiased estimates of intrinsic DNA cyclizability for 
each sequence in the original training dataset. We then trained a new model on 
the unbiased data with architecture identical to that of DNAcycP, named 
DNAcycP2. More details on this process can be found in the following paper: 
(CITATION).

TODO: fill in citation above

Previously, the measurement bias introduced by the location of the biotin 
tether was not adequately accounted for. By employing data augmentation and 
smoothing with a moving average approach over the length of 1 full helical 
repeat at 1bp resolution in the genome, we can remove this bias while still 
maintaining high resolution, accurate estimates of intrinsic cyclizability.

## Available formats of DNAcycP2 and DNAcycP

DNAcycP2 is available in three formats: A web server available at 
http://DNAcycP.stats.northwestern.edu for real-time prediction and 
visualization of C-score up to 20K bp, a standalone Python package avilable 
for free download from https://github.com/jipingw/DNAcycP2, and an R package 
available for free download from https://github.com/jipingw/dnacycp2-R.

TODO: update web server - possible selection on server of which model to use?

DNAcycP is still available in its two original formats: A web server available 
at http://DNAcycP.stats.northwestern.edu for real-time prediction and 
visualization of C-score up to 20K bp, and a standalone Python package 
available for free download from https://github.com/jipingw/DNAcycP

## DNAcycP2 required R packages

* `basilisk`
* `reticulate`

## DNAcycP2 functions

DNAcycP2 supports the input sequence in two formats: FASTA format (with 
sequence name line beginning with “>”) or directly as an R object. Unlike in 
the web server version where only one sequence is allowed in input for 
prediction, the R package allows multiple sequences in the same input 
file/object. In particular for the R object format, each sequence (which can be 
of length >= 50bp) in the file is regarded as one input sequence for 
prediction, however the computation is most efficient when every sequence has 
length exactly 50bp.

The two main functions in the package are **cycle** and **cycle_fasta**, both 
of which perform cyclizability prediction through a python subprocess. The main 
difference between the two functions is the input type: **cycle** takes an R 
object as its input, while **cycle_fasta** takes a file path as its input. Both 
take an additional argument `smooth` which determines which model to use in 
making predictions:

* `smooth=TRUE`: DNAcycP2 (the model trained on smoothed, unbiased data -- 
recommended) 
* `smooth=FALSE`: DNAcycP (the model trained on the original data).

The `cycle_fasta` is designed for larger files, so it has added parallelization 
capability. To utilize this capability, specify the number of cores to be 
greater than 1 using the `n_cores` argument (default 1). You can also specify 
the length of the sequence that each core will predict on at a given time using 
the `chunk_length` argument (default 100000).

For reference, on a personal computer (16 Gb RAM, M1 chip with 8-core CPU), 
prediction at full parallelization directly on the yeast genome FASTA file 
completes in 12 minutes, and on the hg38 human genome Chromosome I FASTA file 
in just over 4 hours. In our experience, selection of parallelization 
parameters (`n_cores` and `chunk_length`) has little affect when making 
predictions on a personal computer, but if using the package on a 
high-performance compute cluster, prediction time should decrease as the number 
of cores increases. If you do run into memory issues, we first suggest reducing 
`chunk_length`.

We provide two simple example files with the package to show proper usage:

```{r}
library(dnacycp2)
```


### Prediction on a .fasta file

```{r}
ex1_file <- system.file("data", "ex1.fasta", package = "dnacycp2")
ex1_smooth <- dnacycp2::cycle_fasta(
    ex1_file, smooth=TRUE, n_cores=1, chunk_length=1000
)
ex1_original <- dnacycp2::cycle_fasta(
    ex1_file, smooth=FALSE, n_cores=1, chunk_length=1000
)
```

`cycle_fasta` takes the file path as input (`ex1_file`)

`smooth=TRUE` specifies that DNAcycP2 be used to make predictions

`smooth=FALSE` specifies that DNAcycP be used to make predictions

`n_cores=1` specifies that 1 core will be used (set >1 for parallel computation)

`chunk_length=1000` specifies that predictions will be made on sequences of 
length 1000 at a given time

The output (`ex1_smooth` or `ex1_original`) is a list with names starting with 
"cycle"

For example, `ex1.fasta` contains two sequences with IDs "1" and "2" 
respectively.

Therefore, both `ex1_smooth` and `ex1_original` will be lists of length 2 with 
names `cycle_1` and `cycle_2` for the first and second sequences respectively.

Each item in the list (e.g. `ex1_smooth$cycle_1`) is a data.frame object with 
three columns. The first columns is always `position`. When `smooth=TRUE`, the 
second and third columns are `C0S_norm` and `C0S_unnorm`, and when 
`smooth=FALSE` the second and third columns are `C0_norm` and 
`C0_unnorm`. The predicted C-score for either model is the normalized 
output (`C0S_norm` and `C0_norm`), the predictions from the model 
trained based on the standardized loop-seq score (in the case of DNAcycP) or 
the standardized smoothed intrinsic cyclizability estimate (in the case of 
DNAcycP2) of the Tiling library of Basu et al 2021 (i.e. 0 mean unit variance). 
When predictions are made using the original DNAcycP (`smooth=FALSE`), 
`C0_unnorm` is the predicted C-score recovered to the original scale of 
loop-seq score in the Tiling library data from Basu et el 2021. When 
predictions are made using the updated DNAcycP2 (`smooth=TRUE`), 
`C0S_unnorm` is the predicted C-score recovered to the scale of 
standardized raw cyclizability scores of the Tiling library data. The 
standardized scores provide two advantages. As loop-seq may be subject to a 
library-specific constant, standardized C-score is defined with a unified 
baseline as yeast genome (i.e. 0 mean in yeast genome). Secondly, the C-score 
provides statisitcal significance indicator, i.e. a C-score of 1.96 indicates 
97.5% in the distribution.

One further option is to save the predictions to output files. To do so, supply 
the path and output file basename to the `save_path` argument. For example:

```{r, eval=FALSE}
ex1_smooth <- dnacycp2::cycle_fasta(
    ex1_file,
    smooth=TRUE,
    n_cores=2,
    chunk_length=1000, 
    save_path="ex1_smooth"
)
```

This will execute the same predictions as previously, and additionally save two 
files named 'ex1_smooth_cycle_1.txt' and 'ex1_smooth_cycle_2.txt' (one file for 
each sequence) to the current working directory.

### Prediction on a list/vector of sequences


```{r}
ex2_file <- system.file("data", "ex2.txt", package = "dnacycp2")
ex2 <- read.csv(ex2_file, header = FALSE)
ex2_smooth <- dnacycp2::cycle(ex2$V1, smooth=TRUE)
ex2_original <- dnacycp2::cycle(ex2$V1, smooth=FALSE)
```

`cycle` takes the sequences themselves as input, so we first read the file 
(`ex2_file`) and then provide the sequences as input (`ex2$V1`)

`smooth=TRUE` specifies that DNAcycP2 be used to make predictions

`smooth=FALSE` specifies that DNAcycP be used to make predictions

The output (`ex2_smooth` or `ex2_original`) is a list with indices 
corresponding to each sequence from the `sequences` argument 
(here it is `ex2$V1`).

For example, `ex2.txt` contains 100 sequences.
Therefore, both `ex2_smooth` and `ex2_original` will be lists of 
length 100, where each entry in the list corresponds to the 
sequence with its same index.

Each item in the list (e.g. `ex2_smooth[[1]]`) is a data.frame 
object with three columns. The first columns is always `position`. 
When `smooth=TRUE`, the second and third columns are `C0S_norm` and 
`C0S_unnorm`, and when `smooth=FALSE` the second and third columns are 
`C0_norm` and `C0_unnorm`. The predicted C-score for either model is 
the normalized output (`C0S_norm` and `C0_norm`), the predictions from 
the model trained based on the standardized loop-seq score (in the case of 
DNAcycP) or the standardized smoothed intrinsic cyclizability estimate (in 
the case of DNAcycP2) of the Tiling library of Basu et al 2021 (i.e. 0 
mean unit variance). When predictions are made using the original DNAcycP 
(`smooth=FALSE`), `C0_unnorm` is the predicted C-score recovered to the 
original scale of loop-seq score in the Tiling library data from Basu et 
el 2021. When predictions are made using the updated DNAcycP2 
(`smooth=TRUE`), `C0S_unnorm` is the predicted C-score recovered to the 
scale of standardized raw cyclizability scores of the Tiling library data. 
The standardized scores provide two advantages. As loop-seq may be subject 
to a library-specific constant, standardized C-score is defined with a 
unified baseline as yeast genome (i.e. 0 mean in yeast genome). Secondly, 
the C-score provides statisitcal significance indicator, i.e. a C-score of 
1.96 indicates 97.5% in the distribution.

If every sequence has length exactly 50bp (recommended), the computation 
will perform significantly faster.

Similar to the .fasta case, there is an option to save the predictions to 
output files. To do so, supply the path and output file basename to the 
`save_path` argument. For example:

```{r, eval=FALSE}
ex2_smooth <- dnacycp2::cycle(
    ex2$V1, 
    smooth=TRUE, 
    save_path="ex2_smooth"
)
```

This will execute the same predictions as previously, and additionally save two 
files named 'ex2_smooth_C0S_norm.txt' and 'ex2_smooth_C0S_unnorm.txt' to the 
current working directory. For consistency with the Python pacakge, ***it is 
important to note that these files have a different format than the function 
output.*** Namely, rather than writing a single file for every sequence, the 
function always writes two files (regardless of the number of sequences), one 
containing normalized predictions for every sequence (ending in 'C0S_norm.txt' 
or 'C0_norm.txt') and the other containing unnormalized predictions for every 
sequence (ending in 'C0S_unnorm.txt' or 'C0_unnorm.txt'). C-scores in each line 
correspond to the sequence from the `sequences` input in the same order.

For any input sequence, DNAcycP2 predicts the C-score for every 50 bp. 
Regardless of the input sequence format the first C-score in the output file 
corresponds to the sequence from position 1-50, second for 2-51 and so forth.

### Example 3 (Single Sequence):

If you want the predict C-scores for a single sequence, you can follow 
the same protocol as Example 1 or 2, depending on the input format. We 
have included two example files representing the same 1000bp stretch of 
S. Cerevisiae sacCer3 Chromosome I (1:1000) in .fasta and .txt format.

First, we will consider the .fasta format:

```r
ex3_fasta_file <- system.file(
    "data", "ex3_single_seq.fasta", package = "dnacycp2"
)
ex3_fasta_smooth <- dnacycp2::cycle_fasta(ex3_fasta_file,smooth=TRUE)
ex3_fasta_original <- dnacycp2::cycle_fasta(ex3_fasta_file,smooth=FALSE)
```

The output (`ex3_fasta_smooth` or `ex3_fasta_original`) is a list with
 1 entry named "cycle_1".

Let's say we are interested only in the smooth (DNAcycP2), normalized
predictions for the subsequence defined by the first 100bp 
(corresponding to subsequences defined by regions [1,50], [2,51],
..., and [51-100], or `position`s 25, 26, ..., and 75). We can 
access the outputs for this subsequence using the following command:

```r
ex3_fasta_smooth[[1]][1:51,c("position", "C0S_norm")]
```

Or, equivalently,

```r
ex3_fasta_smooth$cycle_1[1:51,c("position", "C0S_norm")]
```

Next, we will consider the .txt format:

```r
ex3_txt_file <- system.file("data", "ex3_single_seq.txt", package = "dnacycp2")
ex3_txt <- read.csv(ex3_txt_file, header = FALSE)
ex3_txt_smooth <- dnacycp2::cycle(ex3_txt$V1, smooth=TRUE)
ex3_txt_original <- dnacycp2::cycle(ex3_txt$V1, smooth=FALSE)
```

The output (`ex3_txt_smooth` or `ex3_txt_original`) is a list with 1 entry 
(unnamed).

Note, that `ex3_fasta_smooth` and `ex3_txt_smooth` are essentially equivalent. 
The only exceptions are perhaps slight rounding differences that come from the 
computation, and that the list `ex3_fasta_smooth` has named entries ('cycle_1') 
while `ex3_txt_smooth` does not. The same applies for `ex3_fasta_original` and 
`ex3_txt_original`.

Therefore, we can use a similar command to access the outputs for our 
subsequence of interest:

```r
ex3_txt_smooth[[1]][1:51,c("position", "C0S_norm")]
```

## References

* Li, K., Carroll, M., Vafabakhsh, R., Wang, X.A. and Wang, J.-P., DNAcycP: A 
Deep Learning Tool for DNA Cyclizability Prediction, *Nucleic Acids Research*, 
2021
* Basu, A., Bobrovnikov, D.G., Qureshi, Z., Kayikcioglu, T., Ngo, T.T.M., 
Ranjan, A., Eustermann, S., Cieza, B., Morgan, M.T., Hejna, M. et al. (2021) 
Measuring DNA mechanics on the genome scale. Nature, 589, 462-467.

# Session info

```{r sessionInfo}
sessionInfo()
```